Implicit yaddayadda

### 2011-10-11 papers/PLDI2012/implicitization.tex
reorganisation du papier, suite aux remarques de Matthieu

Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}.

\MM{$X^s$ n'est pas defini. A priori, il suffit d'ajouter un truc du style ... ``except that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for another abstract value $X^d$ later in the algorithm'' ici.}

The combination of these two techniques considers an ascending sequence of
subsets of the paths between two points in $P_R$.
The current working subset of paths is stored using a compact representation,
such as binary decision diagrams.

\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :

\begin{algorithm}
  \begin{algorithmic}
    \STATE \textit{Subset} $\gets \emptyset$
    \WHILE{true}
    \STATE Ascending iteration on a \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
    \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined})
    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{algo:combined})
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.

Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
enchainer sur l'algo 2.
}

\subsection{Ascending iterations}
\label{subsec:ascending}

The algorithm for computing an inductive invariant over a subgraph is the Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11}.

In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
is not included in the actual $X_{j}^s$.
Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
an interesting path to focus on. When the formula becomes unsatisfiable, it
means the analysis of $p_i$ is finished.

@@

We distinguish two phases when analysing a point $p_i \in P_R$:
\begin{enumerate}
\item We search for paths starting at point $p_i$, such that these paths are in
$P$, i.e are included in the working subgraph. Each time we find a path, we
update the abstract value of the destination point of the path. This is the
phase explained in \ref{subsec:ascending}.
\item  When the previous phase terminates, there is no more paths in
$P$, starting at $p_i$, 
that make the abstract values of the successors grow (otherwise, the SMT
solver would not have answered ``\emph{unsat}''). However, there may
exist such paths in the entire multigraph, that are not in $P$. The second phase
computes these paths and adds them to $P'$. We also use SMT solving to discover
these new paths, but we slightly change the SMT formula given to the SMT solver:
we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
the paths given by the SMT solver won't be in $P$.
\end{enumerate}

@@

We maintain a set of active control points $A$, containing the points in $P_R$
that still need to be processed. We apply the two previously described phases to
the elements of $A$ until $A$ is empty. When $A$ is empty, it means the
invariant computation of the subgraph is finished. As proposed by
\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
These narrowing iterations allow to recover precision lost by
widening, before taking into account new feasible paths.
Thus, our technique combines both the advantages of \emph{Guided Static
Analysis} and \emph{Path-focusing}.
We then restart the analysis with a new subgraph if needed.

Termination of this algorihm is guaranteed, because:
\begin{itemize}
\item 
the subset of paths $P$ is
strictly increasing at each loop iteration, and bounded by the finite set of
paths in the entire graph. 
\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
which guarantees that $P'$
will eventually be empty after a finite number of loop iterations.
\end{itemize}

@@

Our improvement consists in a slight modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the disjunction we
have to choose.
\MM{Le ``we have to choose'' est un peu vague. Tu veux dire ``which
  abstract value of the disjunction has to be chosen to make the
  invariant grow'' ?}
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma(j,k)$, thus know the index of the disjunct to join with.

>>>>

Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

The combination of these two techniques aims at first discovering a precise
inductive invariant for a subset of paths between two points in $P_R$, 
by the mean of ascending and narrowing iterations. When an
inductive invariant has been found, we add new feasible paths to the subset and
compute an inductive invariant for this new subset, starting with the results
from the previous analysis.
In other words, our technique considers an ascending sequence of
subsets of the paths between two points in $P_R$.
We iterate the operations until we consider the whole program, i.e all the
feasible paths. The result will then be an inductive invariant of the entire program.

\begin{algorithm}
  \begin{algorithmic}
    \STATE \textit{Subset} $\gets \emptyset$
    \WHILE{true}
    \STATE Ascending iteration on \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
    \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined})
    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{algo:combined})
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

%\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
%  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
%
%Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
%
%Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
%enchainer sur l'algo 2.
%}

The current working subset of paths, noted $P$, is stored using a
compact representation, such as binary decision diagrams.
We also maintain a set of active control points $A$, containing the points in $P_R$
that still need to be processed. 
We apply the Path-focused technique\cite{Monniaux_Gonnord_SAS11} to
the elements of $A$ until $A$ is empty. 
We also check which paths become feasible, so that we can include them into the
next subset of paths at the next iteration. These new feasible paths are
temporarily stored in a set $P'$. 
When $A$ is empty, it means the
invariant computation of the subgraph is finished. As proposed by
\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
These narrowing iterations allow to recover precision lost by
widening, before taking into account new feasible paths.
Thus, our technique combines both the advantages of \emph{Guided Static
Analysis} and \emph{Path-focusing}.
We then restart the analysis with a new subgraph if needed, by adding into $P$
the paths in $P'$. The set of active points $A$ is assigned to $A'$, which is
the set of control points that are starting points of a new feasible path in
$P'$.

\subsection{Ascending iterations by Path-focusing}
\label{subsec:ascending}

The algorithm for computing an inductive invariant over a subgraph is the
Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11} with special
treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).

In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
is not included in the actual $X_{j}^s$.
Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
an interesting path to focus on. We loop until the formula becomes unsatisfiable,
meaning that the analysis of $p_i$ is finished.

@@

We distinguish two phases when analysing a point $p_i \in P_R$:
\begin{enumerate}
\item We search for paths starting at point $p_i$, such that these paths are in
$P$, i.e are included in the working subgraph. Each time we find a path, we
update the abstract value of the destination point of the path. This is the
phase explained in \ref{subsec:ascending}.
\item  When the previous phase terminates, there is no more paths in
$P$, starting at $p_i$, 
that make the abstract values of the successors grow (otherwise, the SMT
solver would not have answered ``\emph{unsat}''). However, there may
exist such paths in the entire multigraph, that are not in $P$. The second phase
computes these paths and adds them to $P'$ (line~\ref{alg=computeNewPaths} in
algorithm~\ref{algo:combined}). We also use SMT solving to discover
these new paths, but we slightly change the SMT formula given to the SMT solver:
we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
the paths given by the SMT solver won't be in $P$.
\end{enumerate}

@@

\subsection{Termination}
Termination of this algorihm is guaranteed, because:
\begin{itemize}
\item 
the subset of paths $P$ is
strictly increasing at each loop iteration, and bounded by the finite set of
paths in the entire graph. 
\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
which guarantees that $P'$
will eventually be empty after a finite number of loop iterations.
\end{itemize}

@@

Our improvement consists in a slight modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the disjunction has
to be chosen to make the invariant grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma(j,k)$, thus know the index of the disjunct to join with.
